TrenchR: an R package for transparent environmental and ecological biophysics

### 2022-02-10 Manuscript_full.Rmd
Updates
abby edits and comments

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abby Meyer, Yutaro Sakairi,..."
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

Lauren B. Buckley^1^, Bryan Briones Ortiz^1^, Isaac Caruso^1^, Aji
John^1^, Ofir Levy^2^, Abby Meyer^1^, Yutaro Sakairi^1^,...

@@

1.  Much understanding of responses to climate change and variability relies on the
    assumption that organismal body temperatures are equal to temporally
    averaged air temperatures high above the ground. However, most
    organisms live near the ground and acute exposure to solar radiation and thermal extremes
    can substantially elevate their body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses. The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation. We offer utility functions to aid in estimating the organismal and environonmental parameters needed for biophysical ecology.
3.  TrenchR focuses on relatively simple and modular functions so users can create transparent and flexible biophysical models.
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting. To further this aim, we offer a series of educational modules to introduce the field of biophysical ecology.

@@

Many responses of organisms and ecosystems to climate change are
inconsistent with current predictive models [@maguire2015modeling]. Some
of these predictive shortcomings stem from omitting spatial and temporal
environmental variation and how it interacts with organismal phenotypes
(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
Many analyses assume that organismal body temperatures are equal to air
temperate at weather station height (usually 2m). However, air
temperatures near the ground where most organisms reside can be
considerably warmer and absorption of solar radiation can raise body
temperatures well above air temperatures, resulting in potential
discrepancies of tens of degrees C [@sinclair2016; @sunday2014;
@kearney2009]. Applications are increasingly demonstrating the importance of accounting for microclimate variation [@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing].  Acute thermal stress events are often more relevant to
the physiology and energy balance of organisms than daily, monthly, or
annually aggregated environmental conditions. Yet, daily and seasonal
environmental variation is often neglected [@dillon2016introduction;
@dillon2016life]. We aim to help remedy these shortcomings by
introducing the TrenchR package (<https://github.com/trenchproject/TrenchR>) as a component of the TrEnCh project
(<https:www.trenchproject.com>), which builds computational and
visualization tools to Translate Environmental Change into organismal
responses.

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of the
tools has not kept pace with research on climate change responses.
Adoptions of biophysical approaches are expanding with recognition of
the importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr]. Increasing dissemination of computational tools
is facilitating adoption. The TrenchR package is intended to complement
the NicheMapR package, which includes sophisticated microclimate
[@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models. Although the Fortran source code was recently released, the complexity of the functions can make it difficult to understand and modify the source code. Many of the approaches are similar between the packages, but TrenchR focuses on
simple functions to aid the understanding and transparency of
biophysical approaches. The TrenchR functions allow for modular use and
easy adaptation to a variety of organisms and research questions. Some
energy budget calculation components of biophysical models are also
available in the ThermImage package [@tattersall2017]. Although many of
the TrenchR functions are general, some components of TrenchR are best
suited for ectothermic animals. We have limited inclusion of plant
specific biophysical models since they are the focus of several R
packages [tealeaves: [@muir2019tea]; plantecophys:
[@duursma2015plantecophys]].

@@

The TrenchR package aims to promote transparency and reproducibility and
we welcome contributions and corrections from users. Our package is
built using the devtools methodology
(<https://github.com/hadley/devtools>) in Github. Issues and feature
requests can be contributed in Github
(<https://github.com/trenchproject/TrenchR/>). The package is maintained
and checked using Travis CI as a tool for continuous integration.

@@

**Specific biophysical functions** provide biophysical models that have
been built for specific organisms based on their physical properties,
behavior, environment, and life history. We currently provide published
biophysical models for lizards, salamanders, butterflies, grasshoppers,
limpets, mussels, and snails. Most models predict operative
environmental temperatures, which are the equilibrium body temperatures
of organisms with specified physical properties in a specific
microenvironment and assume no heat exchange via metabolism or
evaporation [@bakken1992measurement].

@@

We introduce the functions in several vignettes. A good place to start is
the *Allometry and conversions* vignette, which provides tools for
preparing data such as estimating additional dimensions of organisms
from measured dimensions. The *Estimating microclimates* vignette
provides resources for estimating the environmental conditions
experienced by organisms. This includes estimating solar radiation and
its components, diurnal variation in temperature and radiation,
temperature and windspeed profiles, and soil temperatures and profiles.
Finally, the core biophysical modelling functions are described in a
tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual
functions and then body temperatures can be solved for using either a
generic energy balance or taxa specific biophysical models.

For additional background on microclimate and biophysical modelling, we
have updated a series of educational modules entitled *Physical
Processes in Ecosystems*. We intend the modules to provide a more
contained and accessible introduction to microclimate and biophysical
modelling than that included in classic textbooks
[@campbell2000introduction; @gates1980biophysical]. The modules provide
less detailed coverage of many of the topics contained in the books. The
series of 14 modules starts with an overview of the calculus and physics
principles underlying the modelling. Thermodynamics and energy budget
modelling are then introduced with detailed examples. Modules address
the climate space concept, operative temperatures, biophysical models
for leaves and sheep, and heat flow in soils. We have revised the
modules (which originated from an NSF training grant in 1979 and lacked
broad distribution) to include R code and utilization of TrenchR
functions. We expanded the original series of tutorials to include
introductions by Michael Kearney to the Microclim environmental data
[@kearney2014microclim] and integrating the Microclim data with
biophysical modelling to examine broad scale climatic limits. The
modules are provided in html form and available for download as a pdf at
the bookdown server
(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
R markdown files are available in Github at
<https://github.com/trenchproject/TrenchRmodules>.

## Implementation

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2020, in Santa Fe, New
Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also
incorporated in the *Using energy balances to estimate body
temperatures* tutorial. We first estimate the energy budget with
component functions and we then use an integrated biophysical model to
estimate operative environmental temperatures. We start by generating
environmental inputs. See Appendix S1 for another example examining a
time series of microclimate data and applying and testing the
biophysical models.

We will use the energy budget to estimate body temperature,
$T_b$, which can depart dramatically from the air temperatures due to
energy exchange with the environment. Heat is gained from absorbing
solar and thermal radiation and from metabolic reactions. Heat is lost
through the organism's emission of radiation and the evaporation of
water. The organism exchanges heat with the surrounding air or water via
convection and with the ground via conduction. The balance of these heat
exchanges (omitting metabolism and evaporation, which are often negligible for ectotherms) can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement].

Let us assume the lizard is in an unshaded location where a weather
station at standard height (2m) indicates that the daily air temperature
varies from a minimum of 10°C to a maximum of 25°C and the wind speed
averages 1 m/s. The soil surface temperature varies from a minimum of
15°C to a maximum of 30°C. We assume that atmospheric transmissivity
$\tau=0.7$ and albedo $\rho=0.6$

@@

We first prepare the environmental data for analysis. We will need
to estimate hourly air and soil temperatures and radiation. We start by
estimating the day of year and the timing of sunrise and sunset:

@@

```{r}
#set up input data as variables
#latitude and longitude (degrees) and elevation (m)
lat=35.69; lon= -105.944; elev=2121 
#minimum and maximum of air and soil temperatures, respectively (C)
Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 
#wind speed (m/s)
V=1 
#assumptions
tau=0.7; rho=0.6 #atmospheric transmissivity and albedo
Tb0=25 #initial assumption of body temperature (C)

doy= day_of_year("2020-06-01", format= "%Y-%m-%d") 
snoon= solar_noon(lon=lon, doy=doy) #estimate solar noon
dayl= daylength(lat=lat, doy=doy) #estimate daylength
tr= snoon-dayl/2 #time of sunrise
ts= snoon+dayl/2 #time of sunset
```

@@

```{r}
#estimate zenith angle (degrees)
psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) 
#convert to radians
psi_rad= degree_to_radian(psi_deg) 
#estimate radiation
Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho)
#Separate into direct, diffuse, and reflected solar radiation
Sdir= Srad[1,]; Sdif= Srad[2,]; Sref= Srad[3,]
```

We can then calculate hourly air and soil temperatures based on daily
minimum and maximum temperatures. We select the sine-exponential model
for air temperature and the sine model for surface temperature:

```{r}
#air temperature (C)
Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2) 
#soil temperature (C)
Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s) 
```

We use microclimate models to scale air temperature and wind speed from
weather station height (2m) to lizard height (0.02m). We assume a
surface roughness of $z_0=0.2$, which corresponds to bare sand and
determines the turbulence of airflow. We implement neutral air
temperature and wind speed profiles as well as an air temperature
profile forced by wind speed.

```{r}
#scale air temperature to lizard height
#z is reference height, z is organism height, z0 is surface roughness
Ta_liz= air_temp_profile_neutral(T_r=Ta, zr=2, z0=0.2, z=0.02, T_s=Ts)
Ta_liz2= air_temp_profile(T_r=Ta, u_r=V, zr=2, z0=0.2, z=0.02, T_s=Ts)

#scale wind speed to lizard height
V_liz= wind_speed_profile_neutral(u_r=V, zr=2, z0=0.2,z=0.02)
```

@@

We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
where $Q_{net}$ is the net energy exchange with the environment (W),
$Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net
thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to
convection (W), $Q_{cond}$ is energy exchange due to conduction (W),
$Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is
the energy emitted due to evaporative water loss (W). We will estimate
each term on the right side of the equation in turn. Estimating
$Q_{abs}$ requires the surface area exposed to radiation and the solar
absorptivity of the animal surface ($a$ proportion). We model a 10g
*Sceloporus* lizard with solar absorptivity $a=0.9$
[@gates1980biophysical]. We will initially assume $T_b=T_a$ to
illustrate the calculations before solving for $T_b$ given the
environmental conditions.

```{r}
mass=10 #mass (g)
a=0.9 #solar absorptivity (%)

#estimate surface area (m^2) from mass (g)
A= sa_from_mass(mass, "lizard")
#estimate projected (silhouette) area as a portion of surface area
psa= sapply(psi_deg, prop_silhouette_area, taxa= "lizard", posture= "prostrate")
#change negative values to zero
psa[psa<0]=0
```

@@

```{r}
Qabs= rep(NA, 24)
for(hr in 1:24) Qabs[hr]= Qradiation_absorbed(a=a, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], S_dir=Sdir[hr], S_dif=Sdif[hr], a_s=rho)
#where

@@

We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We
assume the surface emissivity of lizards, $epsilon_s=0.965$
[@barlett1967].

```{r}
epsilon_s=0.965

Qemit= rep(NA, 24)
for(hr in 1:24) Qemit[hr]=Qemitted_thermal_radiation(epsilon=epsilon_s, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], T_b=Ta_liz[hr]+273.15, T_g=Ts[hr]+273.15, T_a=Ta_liz[hr]+273.15, enclosed=FALSE)

#where psa_dir and psi_ref are the proportions of surface area exposed to direct and reflected solar radiation, respectively.
```

```{r}
#Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
ap= airpressure_from_elev(elev)*1000 #estimate Barometric pressure (pascal)

DRYAIRout= DRYAIR(db=Ta, bp=ap, alt=elev)
K= DRYAIRout$thcond #thermal conductivity (Wm^-2K^-1) 
nu= DRYAIRout$viskin #kinematic viscosity (m2 s-1)

svl=0.006 #approximate snout vent length (m) for Sceloporus, used as the characteristic dimension for heat transfer

#We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.
K= mean(K)
nu=mean(nu)

#Estimate the heat transfer coefficient using an empirical relationship for lizards
H_L=heat_transfer_coefficient(V=V_liz,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6) , taxa="lizard_surface")

#Also illustrate estimations using a spherical approximation and a simplified version of the approximation.
H_L2= heat_transfer_coefficient_approximation(V=V_liz,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6), taxa="lizard")
H_L3= heat_transfer_coefficient_simple(V=V_liz,D=svl)
```

We estimate convective heat exchange between the animal and surrounding
air using the following relationship:
$$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$ where an
enhancement factor, $ef$, multiplier can be incorporated to account for
increases in heat exchange resulting from air turbulence in field
conditions. We implement the function in R assuming that 2/3 of the
lizard's surface area is exchanging heat through convection. We
currently estimate no convection since we are assuming $Tb=Ta$ for
illustration purposes:

```{r}
Qconv= rep(NA, 24)
for(hr in 1:24) Qconv[hr]= Qconvection(T_a= Ta_liz[hr]+273.15, T_b= Ta_liz[hr]+273.15, H=H_L, A=A, proportion=0.67, ef=1.3)
```

We estimate conductive heat flow (W) from the lizard to the surface
assuming conductance through the animal tissue is the rate limiting step
as follows: $$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
We implement the estimate in R assuming that 1/3 of the lizard surface
is in contact with the ground and a skin thickness $d$ of $10^{-6}$:

```{r}
Qcond= rep(NA, 24)
for(hr in 1:24) Qcond[hr]=Qconduction_animal(T_g= Ts[hr]+273.15, T_b=Ta_liz[hr]+273.15, d=10^-6, K=K, A=A, proportion=0.33)
```

```{r}
Qmet= 0; Qevap= 0
```

```{r}
Qnet= Qnet_Gates(Qabs=Qabs, Qemit=Qemit, Qconv=Qconv, Qcond=Qcond, Qmet=Qmet, Qevap=Qevap)
```

We constructed the energy balance assuming $T_b=T_a$ for illustration.
Thus, the energy balance is dominated by heat gain from contact with the
warmer surface. We now use a function based on the energy balance above
to estimate body temperature given the environmental conditions:

```{r}
Te= rep(NA, 24)
for(hr in 1:24) Te[hr]= Tb_Gates(A=A, D=svl, psa_dir=psa[hr], psa_ref=1-psa[hr], psa_air=0.67, psa_g=0.33, T_g=Ts[hr]+273.15, T_a=Ta_liz[hr]+273.15, Qabs=Qabs[hr], epsilon=epsilon_s, H_L=H_L, ef=1.3, K=K)
```

```{r}
Te2= rep(NA, 24)

>>>>

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail Meyer, Yutaro Sakairi,..."
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

Lauren B. Buckley^1^, Bryan Briones Ortiz^1^, Isaac Caruso^1^, Aji
John^1^, Ofir Levy^2^, Abby Meyer^1^, Yutaro Sakairi^1^

@@

1.  Much understanding of responses to climate change and variability
    relies on the assumption that organismal body temperatures are equal
    to temporally averaged air temperatures high above the ground.
    However, most organisms live near the ground and acute exposure to
    solar radiation and thermal extremes can substantially elevate their
    body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses. The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation. We offer utility functions to aid in estimating the
    organismal and environonmental parameters needed for biophysical
    ecology.
3.  TrenchR focuses on simple and modular functions so users
    can create transparent and flexible biophysical models.
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting. To further this aim, we offer a series of educational
    modules to introduce the field of biophysical ecology.

@@

Many responses of organisms and ecosystems to climate change are
inconsistent with current predictive models [@maguire2015modeling]. Some
of these predictive shortcomings stem from omitting spatial and temporal
environmental variation and how it interacts with organismal phenotypes
(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
Many analyses assume that organismal body temperatures are equal to air
temperatures at weather station height (usually 2 meters). However, air
temperatures near the ground where most organisms reside can be
considerably warmer, and absorption of solar radiation can raise body
temperatures well above air temperatures, resulting in potential
discrepancies of tens of degrees Celsius [@sinclair2016; @sunday2014;
@kearney2009]. Applications are increasingly demonstrating the
importance of accounting for microclimate variation
[@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing].
Acute thermal stress events are often more relevant to the physiology
and energy balance of organisms than daily, monthly, or annually
averaged (*Abby: changed from aggregated*) environmental conditions. Yet, daily and seasonal
environmental variation is often neglected [@dillon2016introduction;
@dillon2016life]. We aim to help remedy these shortcomings by
introducing the TrenchR package
(<https://github.com/trenchproject/TrenchR>) as a component of the
TrEnCh project (<https:www.trenchproject.com>), which builds
computational and visualization tools to Translate Environmental Change
into organismal responses.

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of these
tools has not kept pace with research on climate change responses.
Adoption of biophysical approaches is expanding with recognition of
the importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr] and the increased dissemination of computational tools. 
The TrenchR package is intended to complement
the NicheMapR package, which includes sophisticated microclimate
[@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
Although the Fortran source code was recently released, the complexity
of NicheMapR functions can make it difficult to understand and modify the
source code. Many of the approaches are similar between the packages,
but TrenchR provides simple functions to aid the understanding and
transparency of biophysical approaches. The TrenchR functions are
modular and easy adapt to a variety of organisms and research
questions. Although many of the TrenchR functions are general, some
components of TrenchR are best suited for ectothermic animals.

We have limited inclusion of plant specific biophysical models since
they are the focus of several R packages [tealeaves: [@muir2019tea];
plantecophys: [@duursma2015plantecophys]]. Some energy budget
calculation components of biophysical models are also available in the
ThermImage package [@tattersall2017]. Microclimate models, with an
emphasis on describing spatial variation, are also available in the
Microclima package [@maclean2019microclima].

@@

The TrenchR package aims to promote transparency and reproducibility.
We welcome contributions and corrections from users. Our package is
built using the devtools methodology
(<https://github.com/hadley/devtools>) in Github. Issues and feature
requests can be contributed in Github
(<https://github.com/trenchproject/TrenchR/>). The package is maintained
and checked using Travis CI as a tool for continuous integration.

(*Abby: maybe add the link to the reference website? trenchproject.github.io/TrenchR/*)

@@

**Specific biophysical functions** provide biophysical models that have
been built for specific organisms based on their physical properties,
behavior, environment, and life history. We currently provide published
biophysical models for lizards, salamanders, butterflies, grasshoppers,
limpets, mussels, and snails. Most models predict operative
environmental temperatures, which are the equilibrium body temperatures
of organisms with specified physical properties in a specific
microclimate and assume no heat exchange via metabolism or
evaporation [@bakken1992measurement].

@@

We introduce the functions in several vignettes. A good place to start
is the *Allometry and conversions* vignette, which provides tools for
preparing data, such as estimating additional dimensions of organisms
from measured dimensions. The *Estimating microclimates* vignette
provides resources for estimating the environmental conditions
experienced by organisms. This includes estimating solar radiation and
its components, diurnal variation in temperature and radiation,
temperature and wind speed profiles, and soil temperatures and profiles.
Finally, the core biophysical modelling functions are described in a
tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual
functions and then body temperatures can be solved for using either a
generic energy balance or taxa specific biophysical models.

For additional background on microclimate and biophysical modelling, we
have updated a series of educational modules entitled 
*Physical Processes in Ecosystems*. We intend the modules to provide a more
contained and accessible introduction to microclimate and biophysical
modelling than that included in classic textbooks
[@campbell2000introduction; @gates1980biophysical]. The modules provide
less detailed coverage of many of the topics contained in the books. The
series of 14 modules starts with an overview of the calculus and physics
principles underlying the modelling. Thermodynamics and energy budget
modelling are then introduced with detailed examples. Modules address
the climate space concept, operative temperatures, biophysical models
for leaves and sheep, and heat flow in soils. We have revised the
modules (which originated from an NSF training grant in 1979 and lacked
broad distribution) to include R code and utilization of TrenchR
functions. We expanded the original series of tutorials to include
introductions by Michael Kearney to the Microclim environmental data
[@kearney2014microclim] and integrating the Microclim data with
biophysical modelling to examine broad scale climatic limits. The
modules are provided in html form and available for download as a pdf at
the bookdown server
(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
R markdown files are available in Github (<https://github.com/trenchproject/TrenchRmodules>).

## Implementation
(*Abby: I might title this "Example Application" or "Demonstration of Use" or other. I would assume implementation to mean how you implemented the functions overall*)

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2020 in Santa Fe, New
Mexico, USA (35.69, -105.944, elevation: 2121 meters). The example is also
incorporated in the *Using energy balances to estimate body temperatures* 
vignette. We start by generating environmental inputs. Using these inputs,
we estimate the energy budget with component functions. Finally, we use 
an integrated biophysical model to estimate operative environmental temperatures.  
See Appendix S1 for another example examining a
time series of microclimate data and applying and testing the
biophysical models.

We will use the energy budget to estimate body temperature, $T_b$, which
can depart dramatically from the air temperatures due to energy exchange
with the environment. Heat is gained from absorbed solar and thermal
radiation and from metabolic reactions. Heat is lost through the
organism's emission of radiation and the evaporation of water. The
organism exchanges heat with the surrounding air or water via convection
and with the ground via conduction. The balance of these heat exchanges
(omitting metabolism and evaporation, which are often negligible for
ectotherms) can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement].
(*Abby: I'm confused at where the distinction is made between Tb and Te, for example if the energy exchanges are part of Tb or Te or both.*)

Let us assume the lizard is in an unshaded location where a weather
station at standard height (2 meters) reports the daily air temperature
varies from a minimum of 10°C to a maximum of 25°C and the wind speed
averages 1 m/s. The soil surface temperature varies from a minimum of
15°C to a maximum of 30°C. We assume that atmospheric transmissivity
$\tau = 0.7$ and albedo $\rho = 0.6$

@@

We first prepare the environmental data for analysis. We estimate 
hourly air and soil temperatures and radiation. We start by
estimating the day of year and the timing of sunrise and sunset:

@@

```{r}
# Set up input data as variables
lat = 35.69; # Latitude (degrees)
lon = -105.944; # Longitude (degrees)
elev = 2121 # Elevation (meters)

Tmin = 10; # Minimum air temperature (C)
Tmax = 25; # Maximum air temperature (C)
Tmin_s = 15; # Minimum soil temperature (C)
Tmax_s = 30; # Maximum soil temperature (C)
V = 1 # Wind speed (m/s)

# Assumptions
tau = 0.7; # Atmospheric transmissivity
rho = 0.6; # Albedo
Tb0 = 25; # Initial assumption of body temperature (C)

doy = day_of_year("2020-06-01", format= "%Y-%m-%d") # Find Julian Date (DOY)
snoon = solar_noon(lon = lon, doy = doy) # Estimate solar noon
dayl = daylength(lat = lat, doy = doy) # Estimate day length
tr = snoon - dayl / 2 # Time of sunrise
ts = snoon + dayl / 2 # Time of sunset
```

@@

```{r}
# Estimate zenith angle (degrees)
psi_deg = sapply(1:24, FUN = zenith_angle, doy = doy, lat = lat, lon = lon) 
# Convert to radians
psi_rad = degree_to_radian(psi_deg) 
# Estimate radiation
Srad = sapply(psi_rad, FUN = estimate_radiation, doy = doy, tau = tau, elev = elev, rho = rho)

# Separate solar radiation into direct, diffuse, and reflected
Sdir = Srad[1,]; # Direct solar radiation
Sdif = Srad[2,]; # Diffuse solar radiation
Sref = Srad[3,]; # Reflected solar radiation
```

We then calculate hourly air and soil temperatures based on daily
minimum and maximum temperatures. We select the sine-exponential model
for air temperature and the sine model for surface temperature:

```{r}
# Air temperature (˚C)
Ta = sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2) 
# Soil temperature (˚C)
Ts = sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s) 
```

We use microclimate models to scale air temperature and wind speed from
weather station height (2 meters) to lizard height (0.02 meters). We assume a
surface roughness of $z_0 = 0.2$ (*Abby: units?*), which corresponds to bare sand and
determines the turbulence of airflow. We implement a neutral air
temperature profile, an air temperature
profile forced by wind speed, and a neutral wind speed profile.

```{r}
# Scale air temperature to lizard height
# T_r = reference temperature; zr = reference height; z0 is surface roughness; z = organism height; T_s = surface temperature
Ta_liz = air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Neutral air temperature profile
Ta_liz2 = air_temp_profile(T_r = Ta, u_r = V, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Air temperature profile forced by wind speed

# Scale wind speed to lizard height
# u_r = reference wind speed; zr = reference height; z0 is surface roughness; z = organism height
V_liz= wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) # Neutral wind speed profile
```

@@

We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
where $Q_{net}$ is the net energy exchange with the environment (W),
$Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net
thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to
convection (W), $Q_{cond}$ is energy exchange due to conduction (W),
$Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is
the energy emitted due to evaporative water loss (W). We will estimate
each term on the right side of the equation in turn. Estimating
$Q_{abs}$ requires the surface area exposed to radiation and the solar
absorptivity of the animal surface ($a$ proportion). We model a 10 gram
*Sceloporus* lizard with solar absorptivity $a = 0.9$
[@gates1980biophysical]. We will initially assume $T_b = T_a$ to
illustrate the calculations before solving for $T_b$ given the
environmental conditions.

```{r}
mass = 10 # Mass (g)
a = 0.9 # Solar absorptivity (proportion)

# Estimate surface area and sihouette area
A = sa_from_mass(mass, "lizard") # surface area (m^2)
psa = sapply(psi_deg, prop_silhouette_area, taxa = "lizard", posture = "prostrate") # Projected silhouette area (m^2)

# Change negative values to zero
psa[psa < 0] = 0
```

@@

```{r}
Qabs = rep(NA, 24)
for(hour in 1:24) {

  Qabs[hour] = Qradiation_absorbed(a = a, A = A, psa_dir = psa[hour], psa_ref = 1 - psa[hour], S_dir = Sdir[hour], S_dif = Sdif[hour], a_s = rho)

}
```

We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We
assume the surface emissivity of lizards, $epsilon_s = 0.965$
[@barlett1967].

```{r}
epsilon_s = 0.965 # Surface emissivity of lizards

Qemit = rep(NA, 24)
for(hour in 1:24) {

  # psa_dir = proportions of surface area exposed to direct solar radiation
  # psi_ref = proportions of surface area exposed to reflected solar radiation
  # A = surface area (m^2)
  Qemit[hour] = Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = psa[hour], psa_ref = 1-psa[hour], T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, enclosed = FALSE)

}
```

```{r}
ap = airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)

# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
DRYAIRout = DRYAIR(db = Ta, bp = ap, alt = elev)
K = DRYAIRout$thcond # Thermal conductivity (Wm^-2K^-1) 
nu = DRYAIRout$viskin # Kinematic viscosity (m2 s-1)

svl = 0.006 # Approximate snout vent length (meters) for Sceloporus, used as the characteristic dimension for heat transfer

# We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.
K = mean(K) # Average thermal conductivity across the day
nu = mean(nu) # Average kinematic viscosity

# Estimate the heat transfer coefficient using an empirical relationship for lizards
H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxa = "lizard_surface")

# Estimate the heat transfer coefficient using a spherical approximation
H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxa = "lizard")

# Estimate the heat transfer coefficient using a simplified version of the approximation
H_L3= heat_transfer_coefficient_simple(V = V_liz, D = svl, type = "Gates")
```
*Abby: Added type=Gates above to fix error in H_L3, change if Gates is not the equation wanted. Also, maybe give a reason why one might want to use a spherical or simplified version*

We estimate convective heat exchange between the animal and surrounding
air using the following relationship:
$$Q_{conv} = ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$ where an
enhancement factor, $ef$, multiplier can be incorporated to account for
increases in heat exchange resulting from air turbulence in field
conditions. We implement the function in R assuming that 2/3 of the
lizard's surface area is exchanging heat through convection. We
currently estimate no convection since we are assuming $Tb = Ta$ for
illustration purposes:

```{r}
Qconv = rep(NA, 24)
for(hour in 1:24) {

  Qconv[hour] = Qconvection(T_a = Ta_liz[hour] + 273.15, T_b = Ta_liz[hour] + 273.15, H = H_L, A = A, proportion = 0.67, ef = 1.3)

}
```

We estimate conductive heat flow (W) from the lizard to the surface
assuming conductance through the animal tissue is the rate limiting step
as follows: $$Q_{cond} = A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
We implement the estimate in R assuming that 1/3 of the lizard surface
is in contact with the ground and a skin thickness $d = 10^{-6}$:

```{r}
Qcond = rep(NA, 24)
for(hr in 1:24) {

  Qcond[hr] = Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 10^-6, K = K, A = A, proportion = 0.33)

}
```

```{r}
Qmet = 0; 
Qevap= 0;
```

```{r}
Qnet = Qnet_Gates(Qabs = Qabs, Qemit = Qemit, Qconv = Qconv, Qcond = Qcond, Qmet = Qmet, Qevap = Qevap)
```

We constructed the energy balance assuming $T_b = T_a$ for illustration.
Thus, the energy balance is dominated by heat gain from contact with the
warmer surface. We now use a function based on the energy balance above
to estimate body temperature given the environmental conditions:

```{r}
Te = rep(NA, 24)
for(hour in 1:24) {

  Te[hour]= Tb_Gates(A = A, D = svl, psa_dir = psa[hour], psa_ref = 1 - psa[hour], psa_air = 0.67, psa_g = 0.33, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K)

}
```

```{r}
Te2 = rep(NA, 24)
for(hr in 1:24) {

  # S is solar radiation flux (W m^-2), so we divide by surface area, A
  Te2[hr] = Tb_CampbellNorman(T_a = Ta_liz[hr] + 273.15, T_g = Ts[hr] + 273.15, S = Qabs[hr] / A, alpha_L = 0.96, epsilon = epsilon_s, c_p = 29.3, D = svl, V = V_liz)

}

@@
